---
title: "DSAN 5100 Hypothesis Testing"
author: "Emily Mitchum"
format: 
  html:
    embed-resources: true
    code-fold: false
editor: visual
toc: true
---

# Imports

```{r}
library(stats)
library(ggplot2)
library(dplyr)
```

# Read in Clean Data

```{r}
earnings_df <-read.csv('/Users/Emi/Library/CloudStorage/GoogleDrive-emilymitchum265@gmail.com/Shared drives/DSAN/DSAN 5100/Final Project/vocational_school_employment_outcomes/Cleaned data/Cleaned_quarterly_earnings_data.csv')
head(earnings_df)

tuition_df <- read.csv('/Users/Emi/Library/CloudStorage/GoogleDrive-emilymitchum265@gmail.com/Shared drives/DSAN/DSAN 5100/Final Project/vocational_school_employment_outcomes/Cleaned data/Cleaned_tuition_data.csv')
head(tuition_df)
```

# ANOVA on Quarterly Median Weekly Earnings Data

```{r}
# Correct data types before ANOVA
earnings_df <- earnings_df %>%
  mutate(
    year = factor(format(as.Date(date), "%Y")),
    education_level = factor(education_level)
  )

tuition_df <- tuition_df |> mutate(year = factor(format(as.Date(year_parsed), "%Y")))
```

## Visual Analysis of Earnings by Education Level

```{r}
library(ggplot2)

ggplot(earnings_df, aes(
  x = education_level,
  y = median_weekly_earnings,
  fill = education_level
)) +
  geom_boxplot(alpha = 0.85, outlier.color = "gray40") +
  scale_fill_brewer(palette = "Set2") +   # cleaner, professional palette
  labs(
    title = "Median Weekly Earnings by Education Level",
    x = "Education Level",
    y = "Median Weekly Earnings"
  ) +
  theme_minimal(base_size = 16) +         # larger base text for slides
  theme(
    text = element_text(family = "Helvetica", face = "bold"),
    plot.title = element_text(size = 22, face = "bold", hjust = 0.5),
    axis.title.x = element_text(size = 18, face = "bold"),
    axis.title.y = element_text(size = 18, face = "bold"),
    axis.text.x = element_text(size = 14, face = "bold", angle = 20, hjust = 1),
    axis.text.y = element_text(size = 14, face = "bold"),
    legend.position = "none",             # removes redundant legend
    panel.grid.major.x = element_blank()  # cleaner slide aesthetic
  )

```

## One-Way ANOVA (Ignoring Time)

```{r}
oneway <- aov(median_weekly_earnings ~ education_level, data = earnings_df)
summary(oneway)
```

The one-way ANOVA test revealed a highly significant effect of education level on median weekly earnings. This indicates that mean earnings differ substantially across the three education groups.

## **Two-Way ANOVA (Analyzing Effect of Education Level & Time)**

```{r}
model_two_way <- aov(median_weekly_earnings ~ education_level * year, data = earnings_df)

summary(model_two_way)
```

Bachelor’s degree holders consistently earn much more than those with some college/associate degrees, who in turn earn more than those with only a high school diploma.

```{r}
# map <- c("Non-Degree Granting" = "Vocational", "Degree Granting" = "Traditional Bachelor's")

limited_earnings_df <- earnings_df[earnings_df$education_level != "high_school",] 

limited_earnings_df <- limited_earnings_df |> 
  mutate(degree_category = case_when(
    education_level == "ba_only" ~ "Bachelor's",education_level == "some_college_assoc" ~ "Vocational",
    TRUE ~ "NA"
  ))

tuition_df <- tuition_df |> 
  mutate(degree_category = case_when(
    degree_granting == "Degree Granting" ~ "Bachelor's",
    degree_granting == "Non-Degree Granting" ~ "Vocational",
    TRUE ~ "NA"
  ))

tuition_df
```

Join Dataframes together to assess ROI
```{r}
temp_earnings_df <- limited_earnings_df |> select(c("median_weekly_earnings","year","degree_category"))
temp_tuition_df <- tuition_df |> select(c("average_tuition","year","degree_category"))

merged_df <- merge(temp_earnings_df, temp_tuition_df, by = c("degree_category","year"))

roi_df <- merged_df |> mutate( ROI = median_weekly_earnings / average_tuition )

roi_yearly <- roi_df %>%
  group_by(degree_category, year) %>%
  summarize(ROI = mean(ROI), .groups = "drop") %>%
  arrange(degree_category, year)

plot <- plot_ly(
  roi_yearly,
  x = ~year,
  y = ~ROI,
  color = ~degree_category,
  colors = c("Bachelor's" = "#E64B35", "Vocational" = "#4DBBD5"),
  type = 'scatter',
  mode = 'lines+markers'
) %>%
  layout(
    title = "Return on Investment (ROI) Over Time",
    xaxis = list(title = "Year"),
    yaxis = list(title = "ROI (Median Weekly Earnings / Tuition)"),
    legend = list(title = list(text = "<b>Degree Type</b>"))
  )

htmlwidgets::saveWidget(plot, "roi_plot.html")

roi_yearly
```
